New cardinality estimation algorithms for HyperLogLog sketches

### 2016-06-18 paper/paper.tex
some improvements

\subsection{Depoissonization}
\label{sec:depoissonization}
Due to the simpler probability mass function, it is easier to find an estimator $\symPoissonRateEstimate = \symPoissonRateEstimate(\vec{\symRegValVariate})$ for the Poisson rate $\symPoissonRate$ rather than for the cardinality $\symCardinality$ in the fixed-size model \eqref{equ:multinomialProbabilityMass}. Here $\vec{\symRegValVariate}$ denotes the vector of all observed register values. $\vec{\symRegValVariate}$ is distributed according to \eqref{equ:multinomialProbabilityMass}. Depoissonization \cite{Jacquet1998} finally allows to translate the estimates back to the fixed-size model. Assume we have found an unbiased estimator for the Poisson rate
\begin{equation}
\symExpectation(\symPoissonRateEstimate\vert\symPoissonRate) = \symPoissonRate
\quad
\text{for all $\symPoissonRate\geq 0$}.
\end{equation}
We know from \eqref{equ:poissonization} 
\begin{equation}
\symExpectation(\symPoissonRateEstimate\vert\symPoissonRate) = 
\sum_{\symCardinality=0}^\infty \symExpectation(\symPoissonRateEstimate\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
\end{equation}
and therefore
\begin{equation}
\sum_{\symCardinality=0}^\infty \symExpectation(\symPoissonRateEstimate\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
=
\symPoissonRate
\end{equation}
holds for all $\symPoissonRate\geq0$. The inverse Poisson transform gives
\begin{equation}
\symExpectation(\symPoissonRateEstimate\vert\symCardinality) = \symCardinality.
\end{equation}
Therefore, the unbiased estimator $\symPoissonRateEstimate$ conditioned on $\symCardinality$ is also an unbiased estimator for $\symCardinality$. Hence, it makes sense to use $\symPoissonRateEstimate$ directly as estimate for the cardinality $\symCardinalityEstimate := \symPoissonRateEstimate$. As simulation results will show later, the Poisson approximation works well over the entire cardinality range, even for estimators that are not exactly unbiased.

\section{Original cardinality estimation method}
\label{sec:cardinality_estimation}
The original cardinality estimate \cite{Flajolet2007} is based on the idea that approximately $\symNumReg  2^{\symRegVal_\symIndexJ}$ distinct element insertions are needed until the value of register $\symIndexJ$ reaches $\symRegVal_\symIndexJ$. Given that, a cardinality estimate can be obtained by calculating the average over the values $\lbrace \symNumReg 2^{\symRegVal_1},\ldots,\symNumReg2^{\symRegVal_\symNumReg}\rbrace$.

In the history of the HyperLogLog algorithm different averaging techniques have been proposed. First, there was the LogLog algorithm using the geometric mean and the SuperLogLog algorithm that enhanced the estimate by truncating the largest register values before applying the geometric mean \cite{Durand2003}. Finally, the harmonic mean was found to give even better estimates as it is inherently less sensitive to outliers. The result is the so-called raw estimator and given by
\begin{equation}
\label{equ:intermediate_estimate}
\symCardinalityOriginalEstimate
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexJ=1}^{\symNumReg}2^{-\symRegVal_\symIndexJ}}
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symRegVal=0}^{\symRegRange+1}\symCount_\symRegVal 2^{-\symRegVal}}
\end{equation}
where $\symAlpha_\symNumReg$  is a correction factor that is fixed for a given number of registers $\symNumReg$ and which was derived to be \cite{Flajolet2007}
\begin{equation}
\symAlpha_\symNumReg := \left(
\symNumReg
\int_0^\infty
\left(
\log_2\left(
\frac{2+u}{1+u}
\right)
\right)^\symNumReg
du
\right)^{-1}.
\end{equation}
Numerical approximations of $\symAlpha_\symNumReg$ for various values of $\symNumReg$ have been listed in \cite{Flajolet2007}. These approximations are used in many HyperLogLog implementations. However, since the published constants have been rounded to 4 significant digits, these approximations even introduce some bias for very high precisions. For HyperLogLog sketches that are used in practice with 256 or more registers ($\symPrecision\geq 8$), it is completely sufficient to use 
\begin{equation}
\label{equ:alpha_inf}
\symAlpha_\infty := \lim_{\symNumReg\rightarrow\infty} \symAlpha_\symNumReg = \frac{1}{2\log 2} \approx 0.7213475,
\end{equation}
as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:intermediate_estimate}, because the additional bias is negligible if compared to the estimation error.

\cref{fig:raw_estimate} shows the distribution of the relative error for the raw estimate as function of the cardinality. The chart is based on 10\,000 randomly generated HyperLogLog sketches. Obviously, the raw estimator is biased for small and large cardinalities and fails to return accurate estimates. Therefore, to cover the entire range of cardinalities, corrections for small and large cardinalities have been proposed, respectively.

@@

For large cardinalities in the order of $2^{\symPrecision+\symRegRange}$, for which a lot of registers are already in a saturated state meaning that they have reached the maximum possible value $\symRegRange + 1$, the raw estimator underestimates the cardinalities. For the 32-bit hash value case $(\symPrecision+\symRegRange=32)$ which was considered in \cite{Flajolet2007}, following correction formula was proposed to take these saturated registers into account
\begin{equation}
\label{equ:large_range_estimate}
\symCardinalityEstimate_\text{large}
=
-2^{32}\log\!\left(1-\symCardinalityOriginalEstimate/2^{32}\right).
\end{equation}

The original estimation algorithm as presented in \cite{Flajolet2007} including small and large range corrections is summarized by \cref{alg:estimate_original}. 
\begin{algorithm}
\caption{Original procedure for estimating the cardinality from a  HyperLogLog data structure using 32-bit hash values ($\symPrecision+\symRegRange = 32$) for insertion of data items \cite{Flajolet2007}.}
\label{alg:estimate_original}
\begin{algorithmic}
\Function {EstimateCardinality}{$\vec{\symRegVal}$}
\State $\symNumReg \gets \dim \vec{\symRegVal}$
\State $\symCardinalityOriginalEstimate = \symAlpha_\symNumReg \symNumReg^2\left(\sum_{\symIndexJ=1}^\symNumReg 2^{-\symRegVal_\symIndexJ}\right)^{-1}$
\Comment raw estimate \eqref{equ:intermediate_estimate}
\If{$\symCardinalityOriginalEstimate\leq \frac{5}{2}\symNumReg$}
\State $\symCount_0 = \left|\{\symIndexJ |\symRegVal_\symIndexJ=0\}\right|$
\If{$\symCount_0\neq0$}
\State \Return $\symNumReg \log(\symNumReg / \symCount_0)$ \Comment small range correction \eqref{equ:small_range_estimate}
\Else
\State \Return $\symCardinalityOriginalEstimate$
\EndIf
\ElsIf{$\symCardinalityOriginalEstimate\leq \frac{1}{30}2^{32}$}
\State \Return $\symCardinalityOriginalEstimate$
\Else
\State\Return $-2^{32}\log(1-\symCardinalityOriginalEstimate/2^{32})$\Comment large range correction \eqref{equ:large_range_estimate}
\EndIf
\EndFunction
\end{algorithmic}
\end{algorithm}
The relative estimation error for the original method is shown in \cref{fig:original_estimate}. Unfortunatley, as can be clearly seen, the ranges where the estimation error is small for $\symCardinalityOriginalEstimate$ and $\symCardinalityEstimate_\text{small}$ do not overlap. Therefore, the estimation error is much larger near the transition region. To reduce the estimation error for cardinalities close to this region, it was proposed to correct $\symCardinalityOriginalEstimate$ for bias. Empirically collected bias correction data is either stored as set of interpolation points \cite{Heule2013}, as lookup table \cite{Rhodes2015}, or as best-fitting polynomial \cite{Sanfilippo2014}. However, all these empirical approaches treat the symptom and not the cause.

@@

The large range correction is not satisfying either as it does not reduce the estimation error. Quite the contrary, it even increases the estimation error. However, instead of underestimating the cardinalities, they are now overestimated. A simple approach to avoid the large range correction at all is the use of larger hash values $(\symPrecision+\symRegRange>32)$ which shift the biased region of $\symCardinalityOriginalEstimate$ to higher cardinalities. However, in case $\symRegRange \geq 30$ 6-bit registers as used in \cite{Heule2013} are required to allow representing all values in the range $[0, \symRegRange+1]$. If 64-bit hash values are used as proposed in \cite{Heule2013} $(\symPrecision+\symRegRange=64)$, a correction for large cardinalities can be circumvented. Cardinalities in the order of $2^{64}$ for which saturated registers become a problem never occur in practice.

@@

\subsection{Limitations of the raw estimator}
As we have already seen in \cref{fig:raw_estimate}, the raw estimator does not work very well for small or large cardinalities. The reason for this is that the distribution of register values \eqref{equ:register_value_distribution} is different from \eqref{equ:assumed_register_val_distribution} for which the raw estimator was derived. Since a random variable $\symRegValVariate'$ that obeys \eqref{equ:assumed_register_val_distribution} can be transformed into random variable $\symRegValVariate$ that follows \eqref{equ:register_value_distribution} using the transformation $\symRegValVariate = \min\left(\max\left(\symRegValVariate', 0\right), \symRegRange+1\right)$, the register values of a HyperLogLog sketch can be seen as the result after applying this transformation to a sample that is distributed according to \eqref{equ:assumed_register_val_distribution}. The corresponding multiplicities have  following relationships
\begin{equation}
\label{equ:multiplicity_transformation}
\begin{aligned}
\symCount_0 &= \textstyle\sum_{\symRegVal = -\infty}^{0} \symCount'_\symRegVal, \\
\symCount_\symRegVal &=  \symCount'_\symRegVal, \quad 1\leq\symRegVal\leq\symRegRange,\\
\symCount_{\symRegRange+1} &= \textstyle\sum_{\symRegVal = \symRegRange + 1}^{\infty} \symCount'_\symRegVal.
\end{aligned}
\end{equation}
Obviously, as long as $\symCount_0\approx\symCount'_0$ and  $\symCount'_{\symRegRange+1}\approx\symCount_{\symRegRange+1}$ the set of register values is similar to the originating sample. In case $\symCount_0=0$ and $\symCount_{\symRegRange+1}=0$, which implies $\symCount_\symRegVal=0$ for $\symRegVal\leq0$ and $\symRegVal\geq\symRegRange+1$, we can even be sure that the register values are identical to the originating sample. However, for small and large cardinalities $\symCount_0$ and $\symCount_{\symRegRange+1}$ approach $\symNumReg$, respectively, and the raw estimator needs to be corrected.

\subsection{Corrections to the raw estimator}
Given a HyperLogLog sketch with multiplicity vector $\vec{\symCount}$, we try to find estimates $\hat{\symCount}'_\symRegVal$ for $\symCount'_\symRegVal$ for all $\symRegVal\in\mathbb{Z}$ and use them instead in the estimation formula \eqref{equ:raw_estimate_with_multiplicities}. For $\symRegVal\in[1,\symRegRange]$ where $\symCount_\symRegVal = \symCount'_\symRegVal$ we have the trivial estimators 
\begin{equation}
\hat{\symCount}'_\symRegVal := \symCount_\symRegVal,\quad 1\leq\symRegVal\leq\symRegRange.
\end{equation}
To get estimators for all other $\symRegVal$, we consider their expectations
\begin{equation}
\symExpectation(\symCount'_\symRegVal)
=
\symNumReg e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal}}}\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal}}}\right).
\end{equation}
From \eqref{equ:register_value_distribution} we know that $\symExpectation\left(\frac{\symCount_0}{\symNumReg}\right)=
e^{-\frac{\symPoissonRate}{\symNumReg}}$ and $\symExpectation\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)=
e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}$, and therefore, we can also write
\begin{equation}
\symExpectation(\symCount'_\symRegVal)
=
\symNumReg
\symExpectation\left(\frac{\symCount_0}{\symNumReg}\right)^{2^{-\symRegVal}}
\left(1-\symExpectation\left(\frac{\symCount_0}{\symNumReg}\right)^{2^{-\symRegVal}}\right)
\end{equation}
and
\begin{equation}
\symExpectation(\symCount'_\symRegVal)
=
\symNumReg
\symExpectation\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{\symRegRange-\symRegVal}}
\left(1-\symExpectation\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{\symRegRange-\symRegVal}}\right),
\end{equation}
which motivates us to use
\begin{equation}
\hat{\symCount}'_\symRegVal
:=
\symNumReg\cdot
\left(\frac{\symCount_0}{\symNumReg}\right)^{2^{-\symRegVal}}\cdot
\left(1-\left(\frac{\symCount_0}{\symNumReg}\right)^{2^{-\symRegVal}}\right)
\end{equation}
as estimator for $\symRegVal \leq 0$ and
\begin{equation}
\hat{\symCount}'_\symRegVal
:=
\symNumReg\cdot
\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{\symRegRange-\symRegVal}}
\cdot
\left(1-\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{\symRegRange-\symRegVal}}\right)
\end{equation}
as estimator for $\symRegVal \geq \symRegRange+1$, respectively.

Inserting all these estimators into \eqref{equ:raw_estimate_with_multiplicities} finally gives
\begin{equation}
\label{equ:correctedestimator}
\symPoissonRateEstimate 
= 
\frac{\symAlpha_\infty\symNumReg^2}{\sum_{\symRegVal=-\infty}^\infty \hat{\symCount}'_\symRegVal 2^{-\symRegVal}}
=
\frac{\symAlpha_\infty\symNumReg^2}
{
\left(
\symNumReg\,\symSmallCorrectionFunc(\symCount_0/\symNumReg) + \sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal 2^{-\symRegVal} + \symNumReg\, \symLargeCorrectionFunc(1-\symCount_{\symRegRange+1}/\symNumReg) 2^{-(\symRegRange+1)}
\right)}
\end{equation}
which we call the corrected raw estimator. Here $\symNumReg\,\symSmallCorrectionFunc(\symCount_0/\symNumReg)$ and $\symNumReg\,\symLargeCorrectionFunc(1-\symCount_{\symRegRange+1}/\symNumReg)$ are replacements for  $\symCount_0$ and $\symCount_{\symRegRange+1}$ in the raw estimator \eqref{equ:intermediate_estimate}, respectively. The functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ are defined as
\begin{equation}
\symSmallCorrectionFunc(\symX) := 
\symX
+
\sum_{\symRegVal=1}^\infty
\symX^{2^\symRegVal} 2^{\symRegVal-1}
\end{equation}
and
\begin{equation}
\symLargeCorrectionFunc(\symX)
:=
\sum_{\symRegVal=1}^\infty
\symX^{2^{-\symRegVal}}
\cdot
\left(1-\symX^{2^{-\symRegVal}}\right)
\cdot
2^{-(\symRegVal-1)}.
\end{equation}

Using the identity $2\symSmallCorrectionFunc(\symX) + \symLargeCorrectionFunc(\symX) = \symPowerSeriesFunc(\log_2(\log(1/\symX)))/\left(\log(2)\log(1/\symX)\right)$, we get for the linear counting case with $\symRegRange=0$
\begin{equation}
\symPoissonRateEstimate 
= 
\frac{\symNumReg}{\log(2)\left(
2\symSmallCorrectionFunc(\symCount_0/\symNumReg)
+
\symLargeCorrectionFunc(\symCount_0/\symNumReg)
\right)}
=
\frac{\symNumReg\log\!\left(\symNumReg/\symCount_0\right)}{\symPowerSeriesFunc\!\left(\log_2\!\left(\log\!\left(\symNumReg/\symCount_0\right)\right)\right)}
\approx
\symNumReg\log\!\left(\symNumReg/\symCount_0\right),
\end{equation}
which is almost identical to the linear counting estimator used for the small range correction \eqref{equ:small_range_estimate}. Here we used the fact that $\symPowerSeriesFunc$ can be well approximated by 1.
\subsection{Corrected raw estimation algorithm}
The corrected raw estimator \eqref{equ:correctedestimator} leads to a new cardinality estimation algorithm for HyperLogLog sketches. \cref{alg:corrected_raw_estimation} demonstrates the numerical evaluation of the estimator. The values of functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ can be either calculated on-demand or pre-calculated. The possible value range for $\symCount_0$ and $\symCount_{\symRegRange+1}$ is $[0, \symNumReg]$. Therefore, if performance matters, it is possible to precalculate the function values for all possible arguments and keep them in lookup tables of size $\symNumReg+1$.

>>>>

\subsection{Depoissonization}
\label{sec:depoissonization}
Due to the simpler probability mass function, it is easier to find an estimator $\symPoissonRateEstimate = \symPoissonRateEstimate(\vec{\symRegValVariate})$ for the Poisson rate $\symPoissonRate$ rather than for the cardinality $\symCardinality$ in the fixed-size model \eqref{equ:multinomialProbabilityMass}. Here $\vec{\symRegValVariate}$ denotes the vector of all observed register values. Each component of $\vec{\symRegValVariate}$ is distributed according to \eqref{equ:multinomialProbabilityMass}. Depoissonization \cite{Jacquet1998} finally allows to translate the estimates back to the fixed-size model. Assume we have found an unbiased estimator for the Poisson rate
\begin{equation}
\symExpectation(\symPoissonRateEstimate\vert\symPoissonRate) = \symPoissonRate
\quad
\text{for all $\symPoissonRate\geq 0$}.
\end{equation}
We know from \eqref{equ:poissonization} 
\begin{equation}
\symExpectation(\symPoissonRateEstimate\vert\symPoissonRate) = 
\sum_{\symCardinality=0}^\infty \symExpectation(\symPoissonRateEstimate\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
\end{equation}
and therefore
\begin{equation}
\sum_{\symCardinality=0}^\infty \symExpectation(\symPoissonRateEstimate\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
=
\symPoissonRate
\end{equation}
holds for all $\symPoissonRate\geq0$. The inverse Poisson transform gives
\begin{equation}
\symExpectation(\symPoissonRateEstimate\vert\symCardinality) = \symCardinality.
\end{equation}
Therefore, the unbiased estimator $\symPoissonRateEstimate$ conditioned on $\symCardinality$ is also an unbiased estimator for $\symCardinality$. Hence, it makes sense to use $\symPoissonRateEstimate$ directly as estimate for the cardinality $\symCardinalityEstimate := \symPoissonRateEstimate$. As simulation results will show later, the Poisson approximation works well over the entire cardinality range, even for estimators that are not exactly unbiased.

\section{Original cardinality estimation method}
\label{sec:cardinality_estimation}
The original cardinality estimate \cite{Flajolet2007} is based on the idea that approximately $\symNumReg  2^{\symRegVal_\symIndexJ}$ distinct element insertions are needed until the value of register $\symIndexJ$ reaches $\symRegVal_\symIndexJ$. Given that, a rough cardinality estimate can be obtained by calculating the average over the values $\lbrace \symNumReg 2^{\symRegVal_1},\ldots,\symNumReg2^{\symRegVal_\symNumReg}\rbrace$.

In the history of the HyperLogLog algorithm different averaging techniques have been proposed. First, there was the LogLog algorithm using the geometric mean and the SuperLogLog algorithm that enhanced the estimate by truncating the largest register values before applying the geometric mean \cite{Durand2003}. Finally, the harmonic mean was found to give even better estimates as it is inherently less sensitive to outliers. The result is the so-called raw estimator and given by
\begin{equation}
\label{equ:intermediate_estimate}
\symCardinalityOriginalEstimate
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexJ=1}^{\symNumReg}2^{-\symRegVal_\symIndexJ}}
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symRegVal=0}^{\symRegRange+1}\symCount_\symRegVal 2^{-\symRegVal}}
\end{equation}
where $\symAlpha_\symNumReg$  is a correction factor that is fixed for a given number of registers $\symNumReg$ and which was derived to be \cite{Flajolet2007}
\begin{equation}
\symAlpha_\symNumReg := \left(
\symNumReg
\int_0^\infty
\left(
\log_2\left(
\frac{2+u}{1+u}
\right)
\right)^\symNumReg
du
\right)^{-1}.
\end{equation}
Numerical approximations of $\symAlpha_\symNumReg$ for various values of $\symNumReg$ have been listed in \cite{Flajolet2007}. These approximations are used in many HyperLogLog implementations. However, since the published constants have been rounded to 4 significant digits, these approximations even introduce some bias for very high precisions $\symPrecision$. For HyperLogLog sketches that are used in practice with 256 or more registers ($\symPrecision\geq 8$), it is completely sufficient to use 
\begin{equation}
\label{equ:alpha_inf}
\symAlpha_\infty := \lim_{\symNumReg\rightarrow\infty} \symAlpha_\symNumReg = \frac{1}{2\log 2} \approx 0.7213475,
\end{equation}
as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:intermediate_estimate}, because the additional bias is negligible compared to the estimation error.

\cref{fig:raw_estimate} shows the distribution of the relative error for the raw estimator as function of the cardinality. The chart is based on 10\,000 randomly generated HyperLogLog sketches. Obviously, the raw estimator is biased for small and large cardinalities and fails to return accurate estimates. Therefore, to cover the entire range of cardinalities, corrections for small and large cardinalities have been proposed, respectively.

@@

For large cardinalities in the order of $2^{\symPrecision+\symRegRange}$, for which a lot of registers are already in a saturated state meaning that they have reached the maximum possible value $\symRegRange + 1$, the raw estimator underestimates the cardinalities. For the 32-bit hash value case $(\symPrecision+\symRegRange=32)$ which was considered in \cite{Flajolet2007}, following correction formula was proposed to take these saturated registers into account
\begin{equation}
\label{equ:large_range_estimate}
\symCardinalityEstimate_\text{large}
=
-2^{32}\log\!\left(1-\symCardinalityOriginalEstimate/2^{32}\right).
\end{equation}
The original estimation algorithm as presented in \cite{Flajolet2007} including small and large range corrections is summarized by \cref{alg:estimate_original}. 
\begin{algorithm}
\caption{Original cardinality estimation algorithm for HyperLogLog sketches that use 32-bit hash values ($\symPrecision+\symRegRange = 32$) for insertion of data items \cite{Flajolet2007}.}
\label{alg:estimate_original}
\begin{algorithmic}
\Function {EstimateCardinality}{$\vec{\symRegVal}$}
\State $\symNumReg \gets \dim \vec{\symRegVal}$
\State $\symCardinalityOriginalEstimate = \symAlpha_\symNumReg \symNumReg^2\left(\sum_{\symIndexJ=1}^\symNumReg 2^{-\symRegVal_\symIndexJ}\right)^{-1}$
\Comment raw estimate \eqref{equ:intermediate_estimate}
\If{$\symCardinalityOriginalEstimate\leq \frac{5}{2}\symNumReg$}
\State $\symCount_0 = \left|\{\symIndexJ |\symRegVal_\symIndexJ=0\}\right|$
\If{$\symCount_0\neq0$}
\State \Return $\symNumReg \log(\symNumReg / \symCount_0)$ \Comment small range correction \eqref{equ:small_range_estimate}
\Else
\State \Return $\symCardinalityOriginalEstimate$
\EndIf
\ElsIf{$\symCardinalityOriginalEstimate\leq \frac{1}{30}2^{32}$}
\State \Return $\symCardinalityOriginalEstimate$
\Else
\State\Return $-2^{32}\log(1-\symCardinalityOriginalEstimate/2^{32})$\Comment large range correction \eqref{equ:large_range_estimate}
\EndIf
\EndFunction
\end{algorithmic}
\end{algorithm}
The relative estimation error for the original method is shown in \cref{fig:original_estimate}. Unfortunatley, as can be clearly seen, the ranges where the estimation error is small for $\symCardinalityOriginalEstimate$ and $\symCardinalityEstimate_\text{small}$ do not overlap. Therefore, the estimation error is much larger near the transition region. To reduce the estimation error for cardinalities close to this region, it was proposed to correct $\symCardinalityOriginalEstimate$ for bias. Empirically collected bias correction data is either stored as set of interpolation points \cite{Heule2013}, as lookup table \cite{Rhodes2015}, or as best-fitting polynomial \cite{Sanfilippo2014}. However, all these empirical approaches treat the symptom and not the cause.

@@

The large range correction is not satisfying either as it does not reduce the estimation error. Quite the contrary, it even increases the estimation error. However, instead of underestimating the cardinalities, they are now overestimated. Another indication for the incorrectness of the proposed large range correction \eqref{equ:large_range_estimate} is the fact that it is not even defined for all possible states. For instance, consider a $(\symPrecision,\symRegRange)$-HyperLogLog sketch with $\symPrecision+\symRegRange=32$ for which all registers are equal to the maximum possible value $\symRegRange+1$. The raw estimate would be $\symCardinalityEstimate_\text{raw} = \symAlpha_\symNumReg 2^{33}$, which is greater than $2^{32}$ and outside of the domain of the large range correction.

A simple approach to avoid the large range correction at all is the use of larger hash values $(\symPrecision+\symRegRange>32)$ which shift the biased region of $\symCardinalityOriginalEstimate$ to higher cardinalities. However, in case $\symRegRange \geq 30$ 6-bit registers as used in \cite{Heule2013} are required to allow representing all values in the range $[0, \symRegRange+1]$. If 64-bit hash values are used as proposed in \cite{Heule2013} $(\symPrecision+\symRegRange=64)$, a correction for large cardinalities can be circumvented. Cardinalities in the order of $2^{64}$ for which saturated registers become a problem never occur in practice.

@@

\subsection{Limitations of the raw estimator}
As we have already seen in \cref{fig:raw_estimate}, the raw estimator does not work very well for small or large cardinalities. The reason is that the distribution of register values \eqref{equ:register_value_distribution} is different from \eqref{equ:assumed_register_val_distribution} for which the raw estimator was derived. Since a random variable $\symRegValVariate'$ that obeys \eqref{equ:assumed_register_val_distribution} can be transformed into random variable $\symRegValVariate$ that follows \eqref{equ:register_value_distribution} using the transformation $\symRegValVariate = \min\left(\max\left(\symRegValVariate', 0\right), \symRegRange+1\right)$, the register values of a HyperLogLog sketch can be seen as the result after applying this transformation to a sample that is distributed according to \eqref{equ:assumed_register_val_distribution}. The corresponding multiplicities have  following relationships
\begin{equation}
\label{equ:multiplicity_transformation}
\begin{aligned}
\symCount_0 &= \textstyle\sum_{\symRegVal = -\infty}^{0} \symCount'_\symRegVal, \\
\symCount_\symRegVal &=  \symCount'_\symRegVal, \quad 1\leq\symRegVal\leq\symRegRange,\\
\symCount_{\symRegRange+1} &= \textstyle\sum_{\symRegVal = \symRegRange + 1}^{\infty} \symCount'_\symRegVal.
\end{aligned}
\end{equation}
Obviously, as long as $\symCount_0\approx\symCount'_0$ and  $\symCount'_{\symRegRange+1}\approx\symCount_{\symRegRange+1}$ the set of register values is similar to the originating sample. In case $\symCount_0=0$ and $\symCount_{\symRegRange+1}=0$, which implies $\symCount_\symRegVal=0$ for $\symRegVal\leq0$ and $\symRegVal\geq\symRegRange+1$, we can even be sure that the register values are identical to the originating sample. However, for small and large cardinalities $\symCount_0$ and $\symCount_{\symRegRange+1}$ approach $\symNumReg$, respectively, and the raw estimator needs to be corrected.

\subsection{Corrections to the raw estimator}
Given a HyperLogLog sketch with multiplicity vector $\vec{\symCount}$, we try to find estimates $\hat{\symCount}'_\symRegVal$ for $\symCount'_\symRegVal$ for all $\symRegVal\in\mathbb{Z}$ and use them instead in the estimation formula \eqref{equ:raw_estimate_with_multiplicities}. For $\symRegVal\in[1,\symRegRange]$ where $\symCount_\symRegVal = \symCount'_\symRegVal$ we have the trivial estimators 
\begin{equation}
\hat{\symCount}'_\symRegVal := \symCount_\symRegVal,\quad 1\leq\symRegVal\leq\symRegRange.
\end{equation}
To get estimators for all other $\symRegVal$, we consider their expectations
\begin{equation}
\symExpectation(\symCount'_\symRegVal)
=
\symNumReg e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal}}}\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal}}}\right).
\end{equation}
From \eqref{equ:register_value_distribution} we know that $\symExpectation\!\left(\symCount_0/\symNumReg\right)=
e^{-\frac{\symPoissonRate}{\symNumReg}}$ and $\symExpectation\!\left(1-\symCount_{\symRegRange+1}/\symNumReg\right)=
e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}$, and therefore, we can also write
\begin{equation}
\symExpectation(\symCount'_\symRegVal)
=
\symNumReg
\left(\symExpectation\!\left(\symCount_0/\symNumReg\right)\right)^{2^{-\symRegVal}}
\left(1-\left(\symExpectation\!\left(\symCount_0/\symNumReg\right)\right)^{2^{-\symRegVal}}\right)
\end{equation}
and
\begin{equation}
\symExpectation(\symCount'_\symRegVal)
=
\symNumReg
\left(\symExpectation\!\left(1-\symCount_{\symRegRange+1}/\symNumReg\right)\right)^{2^{\symRegRange-\symRegVal}}
\left(1-\left(\symExpectation\!\left(1-\symCount_{\symRegRange+1}/\symNumReg\right)\right)^{2^{\symRegRange-\symRegVal}}\right),
\end{equation}
which motivates us to use
\begin{equation}
\hat{\symCount}'_\symRegVal
:=
\symNumReg\cdot
\left(\symCount_0/\symNumReg\right)^{2^{-\symRegVal}}\cdot
\left(1-\left(\symCount_0/\symNumReg\right)^{2^{-\symRegVal}}\right)
\end{equation}
as estimator for $\symRegVal \leq 0$ and
\begin{equation}
\hat{\symCount}'_\symRegVal
:=
\symNumReg\cdot
\left(1-\symCount_{\symRegRange+1}/\symNumReg\right)^{2^{\symRegRange-\symRegVal}}
\cdot
\left(1-\left(1-\symCount_{\symRegRange+1}/\symNumReg\right)^{2^{\symRegRange-\symRegVal}}\right)
\end{equation}
as estimator for $\symRegVal \geq \symRegRange+1$, respectively.

Inserting all these estimators into \eqref{equ:raw_estimate_with_multiplicities} as replacements for $\symCount'_\symRegVal$ finally gives
\begin{equation}
\label{equ:correctedestimator}
\symPoissonRateEstimate 
= 
\frac{\symAlpha_\infty\symNumReg^2}{\sum_{\symRegVal=-\infty}^\infty \hat{\symCount}'_\symRegVal 2^{-\symRegVal}}
=
\frac{\symAlpha_\infty\symNumReg^2}
{
\symNumReg\,\symSmallCorrectionFunc(\symCount_0/\symNumReg) + \sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal 2^{-\symRegVal} + \symNumReg\, \symLargeCorrectionFunc(1-\symCount_{\symRegRange+1}/\symNumReg) 2^{-\symRegRange}
}
\end{equation}
which we call the corrected raw estimator. Here $\symNumReg\,\symSmallCorrectionFunc(\symCount_0/\symNumReg)$ and $2\symNumReg\,\symLargeCorrectionFunc(1-\symCount_{\symRegRange+1}/\symNumReg)$ are replacements for  $\symCount_0$ and $\symCount_{\symRegRange+1}$ in the raw estimator \eqref{equ:intermediate_estimate}, respectively. The functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ are defined as
\begin{equation}
\symSmallCorrectionFunc(\symX) := 
\symX
+
\sum_{\symRegVal=1}^\infty
\symX^{2^\symRegVal} 2^{\symRegVal-1}
\end{equation}
and
\begin{equation}
\symLargeCorrectionFunc(\symX)
:=
\sum_{\symRegVal=1}^\infty
\symX^{2^{-\symRegVal}}
\cdot
\left(1-\symX^{2^{-\symRegVal}}\right)
\cdot
2^{-\symRegVal}.
\end{equation}

Using the identity $\symSmallCorrectionFunc(\symX) + \symLargeCorrectionFunc(\symX) = \symAlpha_\infty\symPowerSeriesFunc(\log_2(\log(\symX^{-1})))/\log(\symX^{-1})$, we get for the linear counting case with $\symRegRange=0$
\begin{equation}
\symPoissonRateEstimate 
= 
\frac{\symAlpha_\infty\symNumReg}{
\symSmallCorrectionFunc(\symCount_0/\symNumReg)
+
\symLargeCorrectionFunc(\symCount_0/\symNumReg)
}
=
\frac{\symNumReg\log\!\left(\symNumReg/\symCount_0\right)}{\symPowerSeriesFunc\!\left(\log_2\!\left(\log\!\left(\symNumReg/\symCount_0\right)\right)\right)}
\approx
\symNumReg\log\!\left(\symNumReg/\symCount_0\right),
\end{equation}
which is almost identical to the linear counting estimator used for the small range correction \eqref{equ:small_range_estimate}. Here we used the fact that $\symPowerSeriesFunc$ can be well approximated by 1.
\subsection{Corrected raw estimation algorithm}
The corrected raw estimator \eqref{equ:correctedestimator} leads to a new cardinality estimation algorithm for HyperLogLog sketches. \cref{alg:corrected_raw_estimation} demonstrates the numerical evaluation of the estimator. The values of functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ can be either calculated on-demand or pre-calculated. The possible value range for $\symCount_0$ and $\symCount_{\symRegRange+1}$ is $[0, \symNumReg]$. Therefore, if performance matters, it is possible to precalculate the function values for all possible arguments and keep them in lookup tables of size $\symNumReg+1$.
